### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
gaol: mu.de free monad tweaks

In this section, we try a more ambitious generic operation. Given a functor,
represented as a tagged description, we
build its free monad.
%
\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}
%
Let us recall this construction in, say, Haskell. Given a
functor \texttt{f}, the free monad over \texttt{f} is given:
%
\begin{code}
data FreeMonad f x
    = Var x
    | Composite (f (FreeMonad f x))
\end{code}
%
Provided \texttt{f} is an instance of \texttt{Functor}, we may
take \texttt{Var} for \texttt{return} and use \texttt{f}'s
\texttt{fmap} to define \texttt{>>=} as substitution.

@@

Our free monad construction takes the functor as a
tagged description, and given a set $\V{X}$ of variables, computes the
tagged description of the free monad pattern functor.
%
\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\pair{\DVar{}}{\V{E}}{}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]
%
We simply add a constructor, $\SYMBDVar$, making its arguments
$\DSigma{\V{X}}{\DUnit}$---just an element of $\V{X}$. 
$\V{E}$ and $\V{D}$ stay put, leaving the other constructors
unchanged. Unfolding the interpretation of this definition, we
find an extended sum, corresponding to the \(\V{X}\D{+}\) in
$\FMFreeMonadD$. Taking the fixpoint ties the knot and we have our
data.

@@

Now we need the operations. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, making variables terms. Meanwhile,
\bind is indeed \emph{substitution}, which we now
implement generically, making use of $\cata$. Let us write the type,
and start filling in the blanks:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}\: 
          \SHED
}\]
%
We are left with implementing the algebra of the catamorphism. Its
role is to catch appearances of $\DVar{\V{x}}$ and replace them by
$\V{\sigma}\: \V{x}$. This corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

>>>>

In this section, we try a more ambitious generic operation. Given a
functor---a signature of operations represented as a tagged
description---we build its free monad, extending the signature with
variables and substitution.
%
\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}
%
Let us recall this construction in, say, Haskell. Given a
functor \texttt{f}, the free monad over \texttt{f} is given
thus:
%
\[
\texttt{data FreeMonad f x = Var x | Op (f (FreeMonad f x))}
\]
%
Provided \texttt{f} is an instance of \texttt{Functor}, we may
take \texttt{Var} for \texttt{return} and use \texttt{f}'s
\texttt{fmap} to define \texttt{>>=} as substitution.

@@

Our construction takes the functor as a
tagged description, and given a set $\V{X}$ of variables, computes the
tagged description of the free monad pattern functor.
%
\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\pair{\DVar{}}{\V{E}}{}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]
%
We simply add a constructor, $\SYMBDVar$, making its arguments
$\DSigma{\V{X}}{\DUnit}$---just an element of $\V{X}$. 
$\V{E}$ and $\V{D}$ stay put, leaving the other constructors
unchanged. Unfolding the interpretation of this definition, we
find an extended sum, corresponding to the \(\V{X}\D{+}\) in
$\FMFreeMonadD$. Taking the fixpoint ties the knot and we have our
data.

@@

Now we need the operations. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, making variables terms. Meanwhile,
\bind is indeed \emph{substitution}, which we now
implement generically, making use of $\cata$. Let us write the type,
and start filling in the blanks:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \mude{(\FreeMonad{\V{D}}{\V{Y}})}) \To \\
         & \mude{(\FreeMonad{\V{D}}{\V{X}})} \To
           \mude{(\FreeMonad{\V{D}}{\V{Y}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          (\mude{(\FreeMonad{\V{D}}{\V{Y}})})\: 
          \SHED
}\]
%
We are left with implementing the algebra of the catamorphism. Its
role is to catch appearances of $\DVar{\V{x}}$ and replace them by
$\V{\sigma}\: \V{x}$. This corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \mude{(\FreeMonad{\V{D}}{\V{Y}})}) \To \\
         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{(\mude{(\FreeMonad{\V{D}}{\V{Y}})})}
           \To  \mude{(\FreeMonad{\V{D}}{\V{Y}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]
